\texttt{Taxonomic supertree construction with}\texttt{\emph{ Incertae sedis}}\texttt{ taxa}

### 2018-05-09 doc/incertae_sedis.tex
char encoding of intermed

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy.
This would allow us
to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or
perhaps outside of all named families.
The simplest approach to
placement involves noting whenever a taxon $B$ is a descendant of a
taxon $A$ on the named synthesis tree but not the taxonomy tree.
For
example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
descendant of taxon $A$ in the synthetic tree 
(Fig.\ref{fig:Handling-incertae-sedis}d), but not the taxonomy 
(Fig.\ref{fig:Handling-incertae-sedis}b).
Thus, we could say that the synthetic tree places
$C$ within $A$. Furthermore, the synthetic tree additionally places
$C$ within $AB$.

A slightly more general case is shown in 
Figure \ref{fig:Broken-incertae-sedis}.
Here taxon $T1$ is an descendant of $R$ in the
synthesis tree, but not the taxonomy.
Thus $T1$ is placed within $R$.
We note that the MRCA of $R$ and $T1$ in
the taxonomy is the root node, and the path to the MRCA from $R$ is
$R\to RS\to root$, while the path to the MRCA from $T1$ is $T1\to T\to
root$.
Here the additional node $RS$ on the path to the MRCA from $R$
indicates that $T1$ is placed within $RS$ as well as $R$.
In contrast,
the additional node $T$ on the path to the MRCA indicates that $T1$ is
a broken \emph{incertae sedis} taxon.
Thus we may consider an
algorithm that, for each taxon $S$ on the synthesis tree, finds the
closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
on the taxonomy.
If not, we compute the paths $A\to A_{1}\to\ldots\to
MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.

@@

@@

\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-nested/tree3-shaded}

}\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/is-nested/tax-shaded}

\subfloat[Synthesis tree (\emph{incertae sedis} ignored)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-no-is-shaded}

}\hfill{}\subfloat[Synthesis tree (\emph{incertae sedis} considered)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-is-shaded}

\caption{\label{fig:Nested-incertae-sedis}Nested incertae sedis taxa.
(a)
Input tree \#1 places CI1 within CG, and C within $A$.
(b) In the
taxonomy, CI is \emph{incertae sedis }within C, while C is \emph{incertae
sedis} under the root.
(c) The synthesis tree when \emph{incertae
sedis} taxa are not considered.
The names AB, A, and CG are lost.
(d) The synthesis tree when \emph{incertae sedis} taxa are considered.
The names AB, A, and CG are regained, as well as the edges leading
to AB and CG.}

\end{figure}
When \emph{incertae sedis} clades are allowed, it is possible for an
\emph{incertae sedis} taxon to be nested within another \emph{incertae
sedis} taxon.
In such a case we must allow the higher-ranked taxon to
be placed within a sibling, while preserving the right of the lower
ranked taxon to be placed within a lower-level sibling (Figure
\ref{fig:Nested-incertae-sedis}).

@@

\begin{figure}
\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}

}\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}

}\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}

\caption{\label{fig:Placing-one-incertae}Placing one incertae sedis group
within another.
Here I is placed within A, and then AI is placed within
A.}
\end{figure}

>>>>

After we have attached taxon names to the synthesis tree, we would
like to interpret the position of these names in terms of placing
\emph{incertae sedis }taxa in a revised taxonomy.
This would allow us
to interpret the synthesis tree as saying that phylogenetic
information has (for example) placed genus A within family B, or
perhaps outside of all named families.
The simplest approach to
placement involves noting whenever a taxon $B$ is a descendant of a
taxon $A$ on the named synthesis tree but not the taxonomy tree.
For
example, in Figure \ref{fig:Handling-incertae-sedis}, taxon $C$ is a
descendant of taxon $A$ in the synthetic tree 
(Fig.\ref{fig:Handling-incertae-sedis}d), but not the taxonomy 
(Fig.\ref{fig:Handling-incertae-sedis}b).
Thus, we could say that the synthetic tree places
$C$ within $A$. 
%Furthermore, the synthetic tree additionally places
% $C$ within $AB$.

In general, after naming the clades in the supertree, the
    taxonomic placement of an \emph{incertae sedis} taxon
    can be detected if the most tip-ward named ancestor of
    the clade in the supertree was not an ancestor in the
    taxonomic tree.
Note that if an \emph{incertae sedis} taxon is broken
    during the supertree algorithm, it can be ambiguous whether
    or not all of the members should be placed in a new
    parent taxon.
See, for example, Figure \ref{fig:Broken-incertae-sedis}.
Here taxon $T1$ is an descendant of $R$ in the
synthesis tree, but not the taxonomy.
Thus $T1$ is placed within $R$; taxon
 $T2$ is similarly placed with in $R$.
$T3$ is not mentioned in either input tree, and the 
    only tree that speaks to its placement is the taxonomic input.
However, the taxon $T$ cannot be monophyletic based on the
     two input trees.
Thus the correct placement of $T3$ is uncertain.

The ambiguity of where to place $T3$ can also be compared to the
    behavior of maximum compatibility tree estimation and parsimony.
The matrix representation context the taxonomic tree's
    character representation of taxon $T$ would include $T1$, $T2$ and $T3$
    all being assigned state 1 and all of the other leaves being assigned 
    state 0.
This character is not compatible with the solution tree. 
The compatability approach would be to conclude that the character encoding 
    that taxon was incorrect and should be ignored.
Since, no other character speaks to the position of $T3$, it would fall to being
    a child of the root of the tree.
In a similar result, the BUILD algorithm would place $T3$ as a child of the root.

In a parsimony view, while the character cannot be mapped onto the solution in
    a homoplasy-free manner, one might hope that the character data that the
    taxonomy was based on was not completely unreliable.
Placing of $T3$ as either sister to $T1$ or $T2$ could explain the character
    representation of taxon $T$ with only two acquisitions of the state 1.
Other placements of $T3$ would require at least three 0 to 1 changes (or a reversion
    to state 0)l
Placing $T3$ at the MRCA 
    of the other members of $T$ corresponds
    to handling of ``rogue'' taxa in an Adams consensus method of the two
    most parsimonious placements of $T3$ under an irreversible parsimony model.

% We note that the MRCA of $R$ and $T1$ in
% the taxonomy is the root node, and the path to the ch from $R$ is
% $R\to RS\to root$, while the path to the MRCA from $T1$ is $T1\to T\to
% root$.
% Here the additional node $RS$ on the path to the MRCA from $R$
% indicates that $T1$ is placed within $RS$ as well as $R$.
% In contrast,
% the additional node $T$ on the path to the MRCA indicates that $T1$ is
% a broken \emph{incertae sedis} taxon.
% Thus we may consider an
% algorithm that, for each taxon $S$ on the synthesis tree, finds the
% closest ancestral taxon $A$, and checks if $A$ is an ancestor of $B$
% on the taxonomy.
% If not, we compute the paths $A\to A_{1}\to\ldots\to
% MRCA(A,B)$ and $B\to B_{1}\to\ldots\to MRCA(A,B)$.

@@

\subsection{Ambiguity of imputed taxonomic characters}
It is tempting to view the matrix representation of the taxonomic tree
    as an encoding of the presence, absence, and uncertainty about
    suites of synapomorphies which allow us to diagnose a taxon.
Clearly by simply representing the taxonomic tree, we will not be
    able to recover the identity or even the number of characters that
    a taxonomist might have used when classifying the organisms in question.
Nevertheless, one might hope that a matrix representation of a taxonomy (such
    as the matrix in Figure \ref{fig:Handling-incertae-sedis-mrp} as a 
    representation of Figure \ref{fig:Handling-incertae-sedis}b) has a unique, uncontroversial 
    association with the taxonomic tree.

The rules for associated leaves in $\mathcal{I}(n)$ with state 1, leaves
     in $\mathcal{E}(n)$, and other leaves with the missing data symbol 
     do allow to encode the taxon $n$ in a taxonomic tree unambiguously.
However, these are not the only set of rules for generating a character matrix
    representation that would cause the taxonomic tree allows the characters to
    be mapped without homoplasy.
For example consider the case of the taxonomy shown in Figure \ref{fig:Placing-one-incertae}a.
Either of the two matrices shown in \ref{fig:Placing-one-incertae-MRP} could encode this
    taxonomy along with its \emph{incertae sedis} interpretation.

\begin{figure}
%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
\subfloat[Taxonomy]{\includegraphics[scale=0.5]{Figures/is-within-is/tax-shaded}
}
\subfloat[Synthesis tree]{\includegraphics[scale=0.5]{Figures/is-within-is/synth1-shaded}
}
\subfloat[Possible intermediate]{\includegraphics[scale=0.5]{Figures/is-within-is/intermed-shaded}
}

\caption{\label{fig:Placing-one-incertae}
One \emph{incertae sedis} group ($W$) being placed within another ($V$), with possible
loss of the taxon $V$.}
\end{figure}

\begin{figure}
%\subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-within-is/tree1\lyxdot tre}}
\subfloat[maximally scored]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0  \\
$V1$ & 1 & ?  \\
$V2$ & 1 & ?  \\
$W$  & 0 & 1  \\
$X$  & 0 & 1 \\
\hline
\end{tabular}
}
\subfloat[less informative]{
    \begin{tabular}{|r|p{2em}|p{2em}|}
\hline
& \multicolumn{2}{|c|}{Character} \\
taxa & $V$ & $WX$  \\
\hline
$U$  & 0 & 0 \\
$V1$ & 1 & ? \\
$V2$ & 1 & ? \\
$W$  & ? & 1 \\
$X$  & 0 & 1 \\
\hline
\end{tabular}
}
\caption{\label{fig:Placing-one-incertae-MRP}
Two possible encodings that would support the $V$ and $WX$ branches and \emph{incertae sedis}
annotation shown in the taxonomy from in Figure \ref{fig:Placing-one-incertae}a.
The representations only vary in the scoring of leaf $W$ for character $V$.}
\end{figure}

@@

% \begin{figure}
% \subfloat[Input tree \#1]{\includegraphics[scale=0.5]{Figures/is-nested/tree3-shaded}

% }\hfill{}\subfloat[Taxonomy tree]{\includegraphics[scale=0.5]{Figures/is-nested/tax-shaded}

% }

% \subfloat[Synthesis tree (\emph{incertae sedis} ignored)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-no-is-shaded}

% }\hfill{}\subfloat[Synthesis tree (\emph{incertae sedis} considered)]{\includegraphics[scale=0.5]{Figures/is-nested/synth3-is-shaded}

% }

% \caption{\label{fig:Nested-incertae-sedis}Nested incertae sedis taxa.
% (a)
% Input tree \#1 places CI1 within CG, and C within $A$.
% (b) In the
% taxonomy, CI is \emph{incertae sedis }within C, while C is \emph{incertae
% sedis} under the root.
% (c) The synthesis tree when \emph{incertae
% sedis} taxa are not considered.
% The names AB, A, and CG are lost.
% (d) The synthesis tree when \emph{incertae sedis} taxa are considered.
% The names AB, A, and CG are regained, as well as the edges leading
% to AB and CG.}
% \end{figure}

When \emph{incertae sedis} clades are allowed, it is possible for an
\emph{incertae sedis} taxon to be nested within another \emph{incertae
sedis} taxon.
In such a case we must allow the higher-ranked taxon to
be placed within a sibling, while preserving the right of the lower
ranked taxon to be placed within a lower-level sibling (Figure
\ref{fig:Nested-incertae-sedis}).

@@
